Investigating the Impact of Macroeconomic Indicators on the FTSE 250 Index Using Machine Learning Methods

### 2024-08-23 tex/paper.tex
introduction revised

on the FTSE 250 Index and the Use of Machine Learning Methods to Forecast the Price of the FTSE 250 Index}
\author{Mateen Khan}
\date{\today}

@@

\begin{abstract}
    This study investigates the impact of the Consumer Price Index (CPI), Bank of England Base Rate, US Dollar (USD) and British Pound Sterling (GBP) exchange rate and the money supply (M3) measure on the price of the FTSE 250 Index (FTSE250).
    In addition, it assesses the effectiveness of radial basis function neural networks as a tool to forecast the price of the FTSE 250 Index based on the same set of macroeconomic variables. This study finds quantitative estimates for the long-term and short-term impacts of the aforementioned macroeconomic indicators on the price of the FTSE 250 Index. It also demonstrates the effectiveness of radial basis function neural networks as a short-term forecasting tool for the price of the FTSE 250 Index.
    \keywords{macroeconomic indicators, FTSE 250, autoregressive distributed lag model, radial basis function neural network, time-series forecasting}
\end{abstract}

@@

\subsection{Context}

The stock market has played a central role in the economic and 
political ascent of many nations ever since its inception (\citeboth{levine2002}).  
However, whenever stock market instabilities have arisen in history they 
have had far-reaching destabilising effects. The Mississippi Bubble was 
an early example of the inherent dangers within stock markets in the form of 
speculative bubbles (\citeboth{mississippibubble}). The role of 
macroeconomic policy in this crash was central, and the effects of the crash 
devastated the French economy through rampant inflation. Nonetheless, stock markets also provide essential benefits by allowing companies to 
raise capital, offering investment opportunities for individuals, and 
supporting economic growth through efficient capital allocation. 
They enhance corporate governance through transparency and accountability, 
facilitate price discovery and market liquidity, and contribute to government 
financing through taxes (\citeboth{shahbaz2016}). Effective and sustainable
macroeconomic policy can greatly aid in this beneficial process. Therefore, it is essential 
for economic policymakers to understand the impact of macroeconomic changes 
on the stock market.

The sensitivity of the stock market to macroeconomic indicators is also of great 
relevance to investors. Macroeconomic indicators provide 
insights into the overall health and direction of the 
economy, which affect investor sentiment and in turn influence market trends.
Changes in the macroeconomy have an effect on both consumers and firms 
which impacts their profitability, because these signals 
help investors anticipate future economic conditions and corporate profits
and consequently investment returns.

The significance of this study is to quantify the effects of macroeconomic 
indicators on the stock market as well as assessing the use of these indicators 
to forecast stock prices over the long-term and short-term, both useful to 
economic policy makers and stock market investors respectively.

Modelling the impact of macroeconomic indicators on stock markets 
typically involves a variety of methods: statistical analysis, 
econometric modelling, scenario planning, and advanced machine learning 
techniques. However, this process contains challenges that require careful 
consideration. In particular, there is a challenge to 
find a subset of macroeconomic indicators which are most relevant in 
stock market dynamics. Additionally, there is the task of
choosing the most appropriate model based on the 
research objectives and data to ensure reliable
and accurate results.

\subsection{Research objectives}

This study aims to explore the relationship between macroeconomic indicators and stock markets 
through modelling. Its primary objective is to address questions such as how specific 
macroeconomic indicators influence stock market prices in major economies and whether machine 
learning models can substitute traditional econometric models in predicting stock market prices. 
Ultimately, it aims to investigate three aspects of modelling the impact of macroeconomic 
indicators on stock markets: 
\begin{enumerate}
    \item To investigate how macroeconomic indicators influence the FTSE250
    \item To investigate the short-term and long-term effects of macroeconomic indicators on the price of the FTSE250 
    \item To investigate whether machine learning techniques can provide accurate short-term and long-term forecasts of the price of the FTSE250 when using a small, coarse-grain dataset.

@@

In order to analyse the dynamic relationship between variables in time series data, this study will employ the Autoregressive Distributed Lag (ARDL) model in conjunction with an Error Correction Model (ECM). By integrating the ARDL model with the ECM, a comprehensive understanding of both the immediate and longer-term relationships between variables is obtained, which is valuable for robust economic analysis and policy-making.
This research also employs radial basis function neural networks to predict the price of the FTSE250. This artificial deep learning network is particularly useful for tasks like function approximation, pattern recognition, and regression when the dataset is small.

\subsection{Organisation of paper}
The remainder of the paper is organised as follows. Section 2 presents a brief review of the literature in this problem space, and explains the rationale behind the selection of macroeconomic variables. Section 3 covers the data utilised in this study. Section 4 outlines the ARDL-ECM model, the methodology, and the results achieved. Section 5 constructs the radial basis function neural network (RBF-NN) model, and presents the results of forecasting the short-term and long-term forecasts for the price of the FTSE250 based on the selected macroeconomic variables. Section 6 concludes the study.

>>>>

on the FTSE 250 Index Using Machine Learning Methods}
\author{Mateen Khan}
\date{\today}

@@

\begin{abstract}
    This study investigates the impact of the Consumer Price Index (CPI), Bank of England Base Rate, US Dollar (USD) and British Pound Sterling (GBP) exchange rate, and the money supply (M3) measure on the price of the FTSE 250 Index (FTSE250).
    In addition, it assesses the effectiveness of radial basis function neural networks as a tool to forecast the price of the FTSE 250 Index based on the same set of macroeconomic variables. This study finds quantitative estimates for the short-term and long-term impacts of the aforementioned macroeconomic indicators on the price of the FTSE 250 Index. It also demonstrates the effectiveness of radial basis function neural networks as a short-term forecasting tool for the price of the FTSE 250 Index.
    \keywords{macroeconomic indicators, FTSE 250, autoregressive distributed lag model, radial basis function neural network, time-series forecasting}
\end{abstract}

@@

The stock market has played a central role in the economic and 
political ascent of many nations ever since its inception (\citeboth{levine2002}).  
However, whenever stock market instabilities have arisen in history they 
have had far-reaching destabilising effects. The Mississippi Bubble was 
an early example of the inherent dangers within stock markets in the form of 
speculative bubbles (\citeboth{mississippibubble}). The role of 
macroeconomic policy in this crash was central, and the effects of the crash 
devastated the economy through rampant inflation. Nonetheless, stock markets also provide essential benefits by allowing companies to 
raise capital, offering investment opportunities for individuals, and 
supporting economic growth through efficient capital allocation (\citeboth{shahbaz2016}). Effective, sustainable
macroeconomic policy can greatly aid in this beneficial process. Therefore, it is essential 
for economic policymakers to understand the impact of macroeconomic changes 
on the stock market.

The sensitivity of the stock market to macroeconomic indicators is also of great 
relevance to investors. Changes in macroeconomic indicators have an effect on both consumer preferences
and business costs which impacts the profitability of firms. This 
leads to discrepancies between the valuation of firms by investors and 
stock prices, which in turn drives trends in the stock 
market price.

The significance of this study is to quantify the effects of macroeconomic 
indicators on the stock market as well as assessing the use of these indicators 
to forecast stock market prices over the long-term and short-term, both useful to 
economic policy makers and stock market investors respectively. 
Modelling of this kind typically involves choosing 
the most appropriate method from a variety: statistical analysis, 
econometric modelling, scenario planning, and machine learning 
techniques. Also, careful economic analysis is required to select
the macroeconomic indicators most relevant in 
the dynamics of a nation's stock market.

The objective of this study is to address how specific 
macroeconomic indicators influence stock market prices in a developed economy, and whether machine 
learning models can substitute traditional econometric models in predicting stock market prices. 
Ultimately, it aims to investigate two aspects of the influence of macroeconomic 
indicators on stock markets: 
\begin{enumerate}
    \item To examine the short-term and long-term effects of macroeconomic indicators on the price of the FTSE250 
    \item To investigate whether machine learning techniques can provide accurate short-term and long-term forecasts of the price of the FTSE250 when using a small, coarse-grain dataset.

@@

In order to analyse the dynamic relationship between variables in 
time series data, this study will employ the Autoregressive Distributed 
Lag (ARDL) model in conjunction with an Error Correction Model (ECM). 
This research also employs a radial basis function neural network (RBFNN) to 
predict the price movements of the FTSE250.

The remainder of the paper is organised as follows. 
Section 2 presents a brief review of the literature in this problem space, 
and explains the rationale behind the selection of macroeconomic variables. 
Section 3 covers the data utilised in this study. Section 4 outlines the 
ARDL-ECM model and the results achieved. 
Section 5 constructs the RBFNN 
model, and presents the forecasting results. Section 6 concludes the study.
